---
title: "Project 1 Vignette: Reading In and Exploring Data via an API"
author: "Kelly Baker"
date: "9/10/2020"
output: 
  rmarkdown::github_document:
    toc: true
    toc_depth: 2
---

```{r setup, warning=FALSE, message=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Introduction
<br>
This vignette walks through how to connect with data through an API. Specifically, we will look at how to access certain endpoints (tables) as well as how join returned datasets, create new variables, and explore the data.
<br>

## Required Packages
<br>
To get started, we'll need to ensure we have the tools installed and loaded into RStudio to read in, manipulate, summarize, and plot the data. Below is an outline of packages that are required to for this project:
<br>

* **dplyr**: for data manipulation
* **ggplot2**: for plotting and graphing
* **readr**: for reading in data
* **tibble**: for creating special data frames with improved printing properties
* **tidyr**: for reshaping data
* **jsonlite**: for parsing API data
* **httr**: for streaming data from an API
* **knitr**: for creating tables with nic printing properties
<br>

We'll read in the packages using the library() function:
```{r packages, echo=TRUE, error=TRUE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(readr)
library(tibble)
library(tidyr)
library(jsonlite)
library(httr)
library(knitr)
```
<br>

# API Function Calls
The following portion of this vignette provides code to access various endpoints of both the NHL Records and NHL Stats API.
<br> 

## Return Franchise data
The function below, called `fetchFran()`, pulls in some simple franchise data for the NHL, including id, first season, last season, most recent team id, common team name, and teams' place name.

```{r fun1, echo=TRUE, error=TRUE, warning=FALSE, message=FALSE}
fetchFran <- function(tab) {
  URL <- "https://records.nhl.com/site/api/franchise"
  franGET <- GET(URL)
  franGET_text <- content(franGET, "text")
  franGET_json <- fromJSON(franGET_text, flatten=TRUE)
  franGET_df <- as.data.frame(franGET_json)
  invisible(franGET_df)
  return(franGET_df)
}
```

## Return Franchise data by Team Name
This function, called `fetchFranName()`, returns franchise data for a specified team. For example, if we wanted to look at the same data as in the above example but only for the Boston Bruins, we would call `fetchFranName(name="Bruins")`. Note: we do not call "Boston Bruins" because the teamCommonName is used as the identifier (this does not include Place Name).
```{r fun2, echo=TRUE, error=TRUE, warning=FALSE, message=FALSE}
fetchFranName <- function(tab, name) {
  URL <- "https://records.nhl.com/site/api/franchise"
  franGET <- GET(URL)
  franGET_text <- content(franGET, "text")
  franGET_json <- fromJSON(franGET_text, flatten=TRUE)
  franGET_df <- as.data.frame(franGET_json)
  invisible(franGET_df)
  return(franGET_df %>% filter(data.teamCommonName==name))
}
```

## Return Franchise data by Team ID
This function allows us to return the same data again, this time using Team ID as the identifier. If we again wanted to pull franchise data for the Boston Bruins, we would use their team id (6 in this dataset) as follows: `fetchFranID(id=6)`
```{r fun3, echo=TRUE, error=TRUE, warning=FALSE, message=FALSE}
fetchFranID <- function(tab, id) {
  URL <- "https://records.nhl.com/site/api/franchise"
  franGET <- GET(URL)
  franGET_text <- content(franGET, "text")
  franGET_json <- fromJSON(franGET_text, flatten=TRUE)
  franGET_df <- as.data.frame(franGET_json)
  invisible(franGET_df)
  return(franGET_df %>% filter(data.id==id))
}
```

## Return Franchise Team Totals data
This function returns key franchise totals for including games played, goals for/against, losses and wins by home or away status,  penalty minutes, overall losses and wins, shutouts, and many more. This is a fairly lengthy dataset, so to see a short sample of what this function returns, we could use the function as follows: `head(fetchFranTotal(n))`
```{r fun4, echo=TRUE, error=TRUE, warning=FALSE, message=FALSE}
fetchFranTotal <- function(tab, ...) {
  URL <- "https://records.nhl.com/site/api/franchise-team-totals"
  franGET <- GET(URL)
  franGET_text <- content(franGET, "text")
  franGET_json <- fromJSON(franGET_text, flatten=TRUE)
  franGET_df <- as.data.frame(franGET_json)
  invisible(franGET_df)
  return(franGET_df)
}
```

## Return Franchise Team Totals by Team Name
This function returns the same data as the function above but for a specified team. For example, to call teams totals for the Montreal Canucks, we code `fetchFranNameTotal(name="Vancouver Canucks")`. (Note: in this function, the name input requires both the place and team name).
```{r fun5, echo=TRUE, error=TRUE, warning=FALSE, message=FALSE}
fetchFranNameTotal <- function(tab, name) {
  URL <- "https://records.nhl.com/site/api/franchise-team-totals"
  franGET <- GET(URL)
  franGET_text <- content(franGET, "text")
  franGET_json <- fromJSON(franGET_text, flatten=TRUE)
  franGET_df <- as.data.frame(franGET_json)
  invisible(franGET_df)
  return(franGET_df %>% filter(data.teamName==name))
}
```

## Return Franchise Team Totals by Franchise ID
This function returns the same information above, but uses franchise ID as the specifying criteria instead of name. To pull team totals for the NY Rangers, we'd use the following code: `fetchFranIDTotal(franID="10")`
```{r fun6, echo=TRUE, error=TRUE, warning=FALSE, message=FALSE}
fetchFranIDTotal <- function(tab, franID) {
  URL <- "https://records.nhl.com/site/api/franchise-team-totals"
  franGET <- GET(URL)
  franGET_text <- content(franGET, "text")
  franGET_json <- fromJSON(franGET_text, flatten=TRUE)
  franGET_df <- as.data.frame(franGET_json)
  invisible(franGET_df)
  return(franGET_df %>% filter(data.franchiseId==franID))
}
```


## Return Franchise Season Records by Franchise ID
The fetchSeasonRecords() function below returns season high data such as season with most goals against, season of most losses, most penalty minutes in a season, home point streak, any many more. As an example, this data can be called for the Detroit Red Wings by specifying ID=12 as follows: `fetchSeasonRecords(ID="12")`
```{r fun7, echo=TRUE, error=TRUE, warning=FALSE, message=FALSE}
fetchSeasonRecords <- function(ID, ...) {
  URL <- "https://records.nhl.com/site/api/franchise-season-records?cayenneExp=franchiseId="
  ID <- ID
  fullURL <- paste0(URL,ID)
  franGET <- GET(fullURL)
  franGET_text <- content(franGET, "text")
  franGET_json <- fromJSON(franGET_text, flatten=TRUE)
  return(franGET_json)
}
```

## Return Goalie Records by Franchise ID
This function returns data on goalies by team ID including name, game type played in, number of games played, number of losses, number of wins, number of seasons played with team, and more. As an example, pulling this data for the Toronto Maple Leafs would be done with the following code: `fetchGoalieRecords(ID="5")`.
```{r fun8, echo=TRUE, error=TRUE, warning=FALSE, message=FALSE}
fetchGoalieRecords <- function(ID, ...) { 
  URL <- "https://records.nhl.com/site/api/franchise-goalie-records?cayenneExp=franchiseId="
  ID <- ID
  fullURL <- paste0(URL,ID)
  franGET <- GET(fullURL)
  franGET_text <- content(franGET, "text")
  franGET_json <- fromJSON(franGET_text, flatten=TRUE)
  return(franGET_json)
}
```

## Return Skater Records by Franchise ID
Skater records, including number of penalty minutes, goals, assists, games played, position played, and more is available by franchise ID. To pull this data for the Maple Leafs, we'd use the following code: `fetchSkaterRecords(ID="5")`
```{r fun9, echo=TRUE, error=TRUE, warning=FALSE, message=FALSE}
fetchSkaterRecords <- function(ID, ...) {
  URL <- "https://records.nhl.com/site/api/franchise-skater-records?cayenneExp=franchiseId="
  ID <- ID
  fullURL <- paste0(URL,ID)
  franGET <- GET(fullURL)
  franGET_text <- content(franGET, "text")
  franGET_json <- fromJSON(franGET_text, flatten=TRUE)
  return(franGET_json)
}
```


## Return Team Stats data
Team statistics can be sliced and diced several different ways. The following functions can be used to call stats data as delineated below:
<br>
* fetchRoster(): roster data for all teams  
* fetchPersonData(): similar to fetchRoster() but with less information.  
* fetchNextGame(): Details of upcoming game for a specified team  
* fetchLastGame(): Details of previous game played for a specified team  
* fetchSeasonStats(): Returns seasons stats for all teams  
* fetchRosterBySeason(): Returns roster for teams by specifying season  
* fetchMultiTeams():Returns stats for multiple specified teams  
* fetchMultiStats(): Returns specified stats for all teams  
<br>

```{r fun10, echo=TRUE, error=TRUE, warning=FALSE, message=FALSE}
fetchRosterLess <- function(ID=NULL, ...) { 
  GET(paste0("https://statsapi.web.nhl.com/api/v1/teams/", ID, "/roster")) 
  %>% content("text") %>% fromJSON(flatten=TRUE)
}


fetchPersonData <- function(...) {GET("https://statsapi.web.nhl.com/api/v1/teams/?expand=person.names") 
  %>% content("text") %>% fromJSON(flatten=TRUE)
}


fetchNextGame <- function(...) {
  nextGameData <- GET("https://statsapi.web.nhl.com/api/v1/teams/?expand=team.schedule.next")
  nextGameText <- content(nextGameData, "text")
  nextGameJSON <- fromJSON(nextGameText, flatten=TRUE)
  return(nextGameJSON[2])
}


fetchLastGame <- function(...) {
  lastGameData <- GET("https://statsapi.web.nhl.com/api/v1/teams/?expand=?expand=team.schedule.previous")
  lastGameText <- content(lastGameData, "text")
  lastGameJSON <- fromJSON(lastGameText, flatten=TRUE)
  return(lastGameJSON[2])
}


fetchSeasonStats <- function() {GET("https://statsapi.web.nhl.com/api/v1/teams/?expand=?expand=team.stats") 
  %>% content("text") %>% fromJSON(flatten=TRUE)
}


fetchRosterBySeason <- function(season=NULL, ...) {
  GET(paste0("https://statsapi.web.nhl.com/api/v1/teams/?expand=team.roster&season=", season)) 
  %>% content("text") %>% fromJSON(flatten=TRUE)
}


fetchMultiTeams <- function(ID=NULL, ID2=NULL, ID3=NULL, ...) {
  GET(paste0("https://statsapi.web.nhl.com/api/v1/teams/?teamId=", ID, ",", ID2, ",", ID3)) 
  %>% content("text") %>% fromJSON(flatten=TRUE)
}


fetchMultiStats <- function(stat=NULL, ...) {
  GET(paste0("https://statsapi.web.nhl.com/api/v1/teams/?stats=stats", stat)) 
  %>% content("text") %>% fromJSON(flatten=TRUE)
}
```
<br>

## Wrapper function
The wrapper function is meant to be applied in order to use any of the above written functions for calling the stats API.
```{r wrap, echo=TRUE, error=TRUE, warning=FALSE, message=FALSE}
fetchAnyStat <- function(ID, ID2, ID3, season, stat, ...) {
  if (ID) { 
   rosterLess <- fetchRosterLess(ID="ID")
  }
  if (ID && ID2 && ID3) {
   multiTeams <- fetchMultiTeams(ID="ID", ID2="ID2", ID3="ID3")  
  }
  if (stat) {
    multiStats <- fetchMultiStats(stat="stat")
  }
  if (season) {
   rosterBySeason <- fetchRosterBySeason(season="season")
  }
  if () {
  personData <- fetchPersonData() 
  nextGame <- fetchNextGame()
  lastGame <- fetchLastGame()
  seasonStats <- fetchSeasonStats()
  }
  
return(list(rosterLess, multiTeams, multiStats, rosterBySeason, personData, nextGame, lastGame, seasonStats))
}
  
```

<br>

# Data Manipulation
<br>

## Joins
<br>
In order to better assess the data, I'd like to combine relevant data sets. For example, I might want to see a team's Skater Records and Goalie Records in one dataset. The code provided below does this for the Montreal Canadiens. Using the fetchGoalieRecords() function, I pull the Canadiens goalie records and save in a new object, canGoalieData. This is a list, so I'll need to convert to a data frame first in order to use in my join. Similarly, I've pulled the Canadien's Skater Records as well (object: canSkateRec). Reviewing the data sets, I see many of the same variable (column) names, but many different ones as well. I've opted to use a full join in order to preserve all unique variables of each data set. This join is saved in a new object called joinedData. 
<br>
I've suppressed the output in markdown, but inspecting this joined data reveals 825 rows and 29 columns. This validates that the join worked as the goalie dataset have 37 rows and the skater data set had 788 rows. 
```{r join, echo=TRUE, error=TRUE, message=FALSE, error=FALSE, results=FALSE}
canGoalieRec <- fetchGoalieRecords(ID="1") 
canGoalieRec <- canGoalieRec %>% as.data.frame()## 37 rows, 29 columns
canadiensSkaterRec <- fetchSkaterRecords(ID=1)
canSkateRec <- as.data.frame(canadiensSkaterRec) ##has 788 rows, 31 vars

joinedData <- full_join(canGoalieRec, canSkateRec)
joinedData
```


## Creating New Variables
<br>
Pulling in franchise data using the fetchFranTotal() function, I filtered the dataset to only include active franchises then put in order of first season. From here, I cut the data into 2 chunks representing franchises more than 50 years old ("Oldest") and team less then 50 years old ("Newest").
```{r var1, echo=TRUE, error=TRUE, warning=FALSE, message=FALSE}
franData <- fetchFranTotal() %>% filter(data.activeFranchise==1) %>% arrange(data.firstSeasonId)
franData <- franData %>% as_tibble() %>% 
  mutate(franAge = ifelse(franData$data.firstSeasonId >= 19701971, "Newest", "Oldest"))
```
<br>
I thought it would also be nice to look at this dataset and be able to quickly ascertain what the win percentage was for each team. In order to accomplish this, I created the variable `winPct` that reflects total wins divided by total games played. This is a particularly interesting variable and reveals much even in table form. After sorting by `winPct`, one can quickly glimpse the winningest franchises in hockey history. 
```{r var2, echo=TRUE, error=TRUE, warning=FALSE, message=FALSE}
franData <- franData %>% as_tibble() %>% mutate(winPct = (data.wins)/(data.gamesPlayed)) %>% arrange(desc(winPct))
```

# Data Exploration & Visualization

## Summarizing Numeric Data
<br>
The tables below provide some basic numeric summarization using data from one of hockey's most storied franchises: the Montreal Canadiens. I've pulled in Skater Records using our fetchSkaterRecords() function, setting ID=1, to get started. Because this data is in list form, I will convert it to a data frame. Next, I've written code to create a basic five number summary of assists and goals by skater position. Because I'll need to generate tables across four positions (listed below), I created a function called summByPosition() to help automate this process.  
*  R - right wing;  
*  C - center;  
*  L - left wing;  
*  D - defensemen  
<br>
The tables generated reveal that right wing(R), left wing(L) and centers(C) score, on average, higher than defensemen (D). This aligns with what I would've expected given the fact that offensive positions are tasked with being a team's primary scorers, while defensive positions are not. I was somewhat surprised to see that, though defense trails offense in assists on average, that it wasn't by much when comparing left wingers to defensemen.
```{r numsumm, echo=TRUE, error=TRUE, warning=FALSE, message=FALSE}
canadiensSkaterRec <- fetchSkaterRecords(ID=1)
canSkateRec <- as.data.frame(canadiensSkaterRec)

summByPosition <- function(position) {
  data <- canSkateRec %>% filter(data.positionCode==position) %>% select(data.assists, data.goals)
  kable(apply(data, 2, summary), col.names=c("Assists", "Goals"), 
    caption = paste0("Historical Summary of Assists & Goals Scored by Position=", position, " for the Montreal Canadiens"))
}

summByPosition("R")
summByPosition("C")
summByPosition("L")
summByPosition("D")
```
<br>

## Summarizing Categorical Data

Working with the franchise records data and the new variable `franAge` created previously, I have created a contingency table below that shows the frequency of franchises by age classification (either Oldest or Newest). Newest franchises are more frequent by almost double that of Oldest franchises, however, that data is somewhat misleading. Each franchise appears twice in the dataset based on type of season. See more below. 
```{r catsumm1, echo=TRUE, error=TRUE, warning=FALSE, message=FALSE}
kable(table(franData$franAge))
```

<br>
Adding a second variable, `gameTypeId`, we get a two way contingency table that shows the frequency of franchises by age status and season type. This table tells us that there are 15 teams classified as "Oldest" and all teams with that classification have had a regular season and a post season in hockey history. However, the Newest franchises appears to have 29 teams classified as such, with one team not ever having had a post-season.
<br>
```{r catsumm2, echo=TRUE, error=TRUE, warning=FALSE, message=FALSE}
kable(table(franData$franAge, franData$data.gameTypeId), col.names = c("Regular Season", "Playoffs"))
```

<br>

# Graphing

## Boxplot: Winning Percentage by Franchise Age
The visual below shows boxplots of winning percentage by franchise age classification. The plots reveal that the newest franchises have greater variability in winning percentage than teams classified "Oldest".
```{r boxplot1, echo=TRUE, error=TRUE, warning=FALSE, message=FALSE}
ggplot(franData, aes(x = franAge, y = winPct)) + geom_boxplot() + geom_jitter(aes(color=franAge)) + 
  ggtitle("Boxplot of Winning Percentage by Franchise Age") + ylab("Winning Percentage") + 
  xlab("Franchise Age Classification")
```
<br>

## Scatterplot: Total Wins v. Road Wins
The overarching purpose of the scatterplot below is to compares total wins and road wins. We can see a vey strong, positive linear association between these two variables, which I expected. Winning games away from home is generally more challenging that in front of a friendly home crowd so I expected as the road wins increased, so too would oveall wins and the success of the franchise. I also a color grouping so that we can see the age of franchises. In this case, the major insight to be drawn from this is that the oldest franchises tend to have the overall most wins -- which again makes sense because older teams have been playing longer and simply have the opportunity to win more games than newer franchises. 
```{r scatter, echo=TRUE, error=TRUE, warning=FALSE, message=FALSE}
ggplot(franData, aes(x = data.wins, y = data.roadWins)) + geom_jitter(aes(color=franAge)) + 
  ggtitle("Scatterplot of Total Wins vs. Road Wins") + ylab("Wins on the Road") + xlab("Total Wins")
```
<br>
## Scatterplot: Total Losses vs. Total Penalty Minutes
The scatterplot below shows total losses compared to total penalty minutes by franchise. This has a moderately strong positive linear association, though not quite as strong as the relationship seen in the previous plot. I expected this correlation given that players in penalty boxes generally give their team a disadvantage by allowing the other team a power play (unless of course the other team is penalized similarly at the exact same time and thus may experience more losses. Time does look to play a role here. We see the oldest franchises again having a higher number of penalty minutes and a higher number of losses, which again make sense because they have simply have longer (i.e more opportunity) to experience those things ove newer franchises. 
```{r scatter2, echo=TRUE, error=TRUE, warning=FALSE, message=FALSE}
ggplot(franData, aes(x = data.losses, y = data.penaltyMinutes, group=franAge)) + geom_jitter(aes(color=franAge)) + 
  ggtitle("Scatterplot of Total Losses vs. Penalty Minutes") + ylab("Total Penalty Minutes") + xlab("Total Losses")
```

<br>
## Histogram: Frequency of Seasons Played by Vancouver Canucks Goalies
The histogram below shows the frequency of seasons played by goalies for the Vancouver Canucks. The distribution is right skewed revealing that goalies most frequently play 1 to 2 seasons.
```{r hist, echo=TRUE, error=TRUE, warning=FALSE, message=FALSE}
vanGoalieData <- fetchGoalieRecords(ID=20)
vanGoalieData <- vanGoalieData[[1]][26]

ggplot(vanGoalieData, aes(x=seasons)) + geom_histogram(bins=10) + 
  ggtitle("Histogram of Seasons Played by Vancouver Canucks Goalies") + xlab("Seasons Played")
```
<br>

## Barplot
Using skater data from the Montreal Canadiens that we looked at earlier, I've developed a barplot to show the count of max goals scored in one game broken down by position. We see defensemen showing high counts for both zero and one goal max per game, which makes sense since it isn't their primary job to score. As the max number of goals per game increases, we see the counts for defensive positions fall off significantly. Counts for offensive positions at all available max goals is similar for zero to two goals max and begins to decline moderatly from there as we look from left to right across the barplot. Centers was the only position to register 6 goals max per game (the largest per game total recorded for the Canadiens).
```{r barplot, echo=TRUE, error=TRUE, warning=FALSE, message=FALSE}
ggplot(canSkateRec, aes(x=data.mostGoalsOneGame)) + geom_bar(aes(fill=data.positionCode), position="dodge") + 
  scale_fill_discrete("Position") + xlab("Max Goals in One Game") + ggtitle("Barplot of Max Goals/Game by Position")
```
